
\section*{LIST OF PUBLICATIONS}
\addcontentsline{toc}{section}{List of publications}
\markboth{LIST OF PUBLICATIONS}{LIST OF PUBLICATIONS}
\label{sec:publications}

This thesis consists of an overview and of the following publications
which are referred to in the text by their Roman numerals.


\leftmargini=6mm {\sloppy
\begin{enumerate}
\item Laura L. Elo, Leo Lahti, Heli Skottman, Minna Kyl{\"a}niemi,
  Riitta Lahesmaa, and Tero Aittokallio.  \newblock Integrating
  probe-level expression changes across generations of Affymetrix
  arrays.  \newblock {\em Nucleic Acids Research}, 33(22):e193, 2005.
\label{PECA}
%  \copyright 2005 The Author. Published by Oxford University PressOxford University Press

\item Leo Lahti, Laura L. Elo, Tero Aittokallio, and Samuel Kaski.
  \newblock Probabilistic analysis of probe reliability in
  differential gene expression studies with short oligonucleotide
  arrays. \newblock {\em IEEE/ACM Transactions on Computational Biology
    and Bioinformatics}, 8(1):217--225, 2011.
\label{RPA}

\item Leo Lahti, Juha E.A. Knuuttila, and Samuel Kaski.
\newblock Global modeling of transcriptional responses in interaction networks.
\newblock {\it Bioinformatics}, 26(21):2713--2720, 2010.
\label{NR}

\item Leo Lahti, Samuel Myllykangas, Sakari Knuutila, and Samuel
  Kaski.  \newblock Dependency detection with similarity constraints.
  \newblock In T{\"u}lay Adali, Jocelyn Chanussot, Christian Jutten,
  and Jan Larsen, editors, {\em Proceedings of the 2009 IEEE
    International Workshop on Machine Learning for Signal Processing
    XIX}, pages 89--94. IEEE, Piscataway, NJ, 2009.  
\label{MLSP}

\item Janne Sinkkonen, Janne Nikkil\"{a}, Leo Lahti, and Samuel Kaski.
  \newblock Associative clustering.  \newblock In Boulicaut, Esposito,
  Giannotti, and Pedreschi (editors), {\em Machine Learning: ECML2004
    (Proceedings of the ECML'04, 15th European Conference on Machine
    Learning)}, Lecture Notes in Computer Science 3201,
  396--406. Springer, Berlin, 2004.  
\label{ECML}

\item
Samuel Kaski, Janne Nikkil{\"a}, Janne Sinkkonen, Leo Lahti, Juha E.A. Knuuttila, and Cristophe Roos.
\newblock Associative clustering for exploring dependencies between functional
  genomics data sets.
\newblock {\em IEEE/ACM Transactions on Computational Biology and
  Bioinformatics: Special Issue on Machine Learning for Bioinformatics -- Part 2}, 2(3):203--216, 2005.
\label{AC}


\end{enumerate}}


